Review Paper Draft

### 2020-06-29 main.tex
Replacing '\cite' with '\citep'

Genetic variants are often correlated with each other, meaning that univariate testing of genetic variants may result in false positives and biased effect estimates. This occurs when the variant being tested is correlated with other variants, so that the effects of these correlated variants are erroneously attributed to the variant in question. \anna{One method for analyzing data in the presence of correlated loci, or linkage disequilibrium (LD), is LD-pruning. In LD-pruning methods, SNPs are sequentially thinned, so that only SNPs correlated with each other below some desired threshold are used in modeling and testing procedures.} LD-pruning limits analysis to SNPs which are representative of a particular region. This may result in causal SNPs being removed from analysis, so that only SNPs in LD with the causal loci can be identified and their effects estimated. In addition, it has been shown that in the worst-case scenario, pruning algorithms may not select any representative SNPs in a particular region of the genome \cite{prive2018efficient}. However, LD-pruning does allow for the use of methods which assume uncorrelated loci such as \anna{BLAHHHHH: \url{https://www.nature.com/articles/srep27644/tables/1}}. In addition, LD-pruning may result in improved predictive accuracy \anna{because why}, which can be of particular benefit to polygenic risk score methods.

@@

The Lasso \cite{tibshirani1996regression} is a widely used penalized regression method. It relies on an assumption of sparsity, which means that coefficient estimation and variable selection can be accomplished simultaneously, and also provides considerable computational advantages, meaning that it scales up well to very large data sets. These features make the Lasso attractive for the analysis of high-dimensional genetic data where identifying a list of variants most highly associated with a particular trait is often of interest.

Lasso penalized regression models typically assume independent observations and uncorrelated errors. In the presence of population structure, some samples are more correlated than others, violating the independent errors assumption. If unaccounted for, this dependency can result in the identification of spurious associations. Though methods to correct for population stratification and cryptic relatedness have been extensively investigated \cite{Amin2007, hoffman2013correcting, price2006principal, Rakitsch2012, bhatnagar2019simultaneous, Sillanpaeae2011}, there remain open and relevant problems \pb{this is vague} in analyzing structured genetic data \cite{lawson2019population, barton2019population}.

One reason that population stratification warrants such concern in genetic studies is that due to ethnic and geographic segregation, population stratification tends to be associated with differing environmental exposures and cultural practices.  In other words, random genetic variation is likely to be associated with non-genetic factors that also influence the trait of interest, thereby confounding the genotype-phenotype relationship and leading to spurious associations and biased estimates of SNP effects. Biased estimates of this nature severely hinder prediction and understanding differences among populations \cite{barton2019population}. Although bias at individual loci may be small, this bias may become magnified when aggregated across thousands of SNPs, as is done when calculating polygenic risk scores \cite{barton2019population}.

In order to clarify and emphasize the importance of this subtle distinction, we present a detailed review of relevant concepts and methods. The remainder of this paper is organized as follows. In section \ref{sec:background} we provide a summary of the causes and consequences of structure in the genetic data of seemingly unrelated individuals and formally define terminology used throughout this paper. We emphasize that confounding due to population structure, frequently cited as a driver of spurious associations \cite{Sillanpaeae2011, sul2018population}, is a function of environmental heterogeneity, where genetic data serves as a proxy for differential environmental exposures. In so doing, we refine the definition of confounding due to population structure in terms of a non-genetic mechanism. In addition, we briefly review historical methods of correcting for population stratification and relatedness. In section \ref{sec:methods} we provide a detailed review of the two most prevalent multivariate methods of adjusting for population stratification and relatedness at present: PCA adjusted lasso penalized regression (PC-Lasso) and lasso penalized multivariate LMMs (LMM-Lasso). We review the statistical details of these methods as well as the consequences of the ways in which they attempt to correct for population structure and environmental heterogeneity. In section \ref{sec:results}, we illustrate the concepts described in section \ref{sec:methods} via simulation studies, where the data-generating mechanism formally differentiates between population structure and environmental heterogeneity. These simulation results are used to illustrate the importance of distinguishing between population structure and heterogeneous environmental exposures in penalized regression methods. In addition, these results identify scenarios where particular methods may be most and least effective in accurately estimating the effect sizes of observed SNPs in the presence of unobserved confounding.

@@

Population structure is a common phenomenon in genetic data. Varying levels of relatedness are almost always present among genetic samples, even in samples of unrelated individuals and seemingly homogeneous populations. For example, European American \cite{campbell2005demonstrating}, Han Chinese \cite{xu2009genomic, chen2009genetic}, and most recently, cohorts within the UK Biobank data \cite{haworth2019apparent}.

@@

Pedigree-based methods may be used to explicitly model recent relatedness if familial relationships are known. However, in this review we focus on methods to account for unobserved relatedness. Throughout this paper, the terms \textbf{population structure} and \textbf{structured population} will be used to describe a sample of individuals in which population stratification, cryptic relatedness, or both are present. Population stratification in particular has been of great concern in genetic studies due to its potential to lead to spurious associations. This phenomenon is commonly described as \textbf{confounding due to population stratification}. However, the mechanism of this phenomenon warrants some discussion. In particular, it is often overlooked that population structure itself does not confound the genotype-phenotype relationship in the classical sense, rather there must exist some non-genetic mechanism by which population stratification affects both the phenotype and genotype \cite{barton2019population, vilhjalmsson2012nature}. \\

@@

An additional consideration in genetic studies that we will briefly mention is that of genetic background. Genetic background describes when SNPs that have a causal relationship with the phenotype of interest are not directly observed, but are correlated with SNPs that are observed. As \cite{vilhjalmsson2012nature} note, any trait that has some genetic basis will be subject to genetic background effects, just as traits that are influenced by the environment may be subject to environmental confounding. Given the importance of both genetic and environmental contributions to many complex phenotypes, both genetic background and environmental confounding are important factors in our ability to identify true causal genetic relationships between SNPs and phenotype. Although there are a number of methods in which to account for genetic background \anna{what are they?}, the methods upon which we will focus address this by conducting a genomewide multivariate analysis.

@@

The genomic control (GC) method is a univariate testing procedure which modifies a test statistic by a deflation factor. This deflation factor attempts to quantify and remove the effect of population structure on the distribution of the appropriate test statistic, and is estimated using loci independent of the phenotype of interest and the loci being tested for association \cite{devlin1999genomic, bacanu2000power, wang2009testing}. However, this deflation factor has been reported to be sensitive to the nature and quantity of the loci used to estimate it \cite{hellwege2017population, marchini2004effects}. Additionally, GC does not correct effect size estimates, so resulting coefficient estimates will be unreliable after GC is applied, even if their corresponding test statistics and p-values have been appropriately adjusted \cite{hellwege2017population}.

Stratified analysis attempts to infer subject membership within a discrete number of nonoverlapping subpopulations. Subsequent analyses are conducted within each subpopulation, and the subpopulation-specific results may then be combined via meta-analysis \cite{pritchard1999use, pritchard2000association}. By partitioning the overall sample size in this manner, the structured association method is particularly vulnerable to a loss of power. Additionally, it relies on the assumption that these subpopulations are distinct and nonoveralapping. This assumption may not accurately reflect the characteristics of a particular sample, such as one consisting of admixed populations, or recent relatedness. Considered in the context of our air quality example, rather than the existence of two subpopulations with exposure to either good or bad air quality, there may also be individuals who fall between these two extremes and are characterized by continuously varying relatedness and a gradient of air quality. By clustering those individuals within in one subpopulation or another we lose information about their fine-grained, subtle relatedness.

Another approach is to create an approximation of population structure using surrogate variables, and to adjust for these as an additional model covariates. Principal component-adjustment (PC-adjustment) methods use the first several principal components of the design matrix as these surrogate variables \cite{price2006principal}. This approach may be used in univariate or multivariate frameworks, and allows for analysis on the entirety of a structured sample.

@@

Among LMM methods that aim to estimate heritability, the most well-known example is genome-wide complex trait analysis (GCTA) \cite{yang2011gcta}. In this framework, all observed SNPs are treated as random effects. These effects are integrated out in order to estimate a genetic variance component, which in turn, is used to quantify the total narrow-sense heritability of a trait \cite{yang2010common}.

LMMs can also be used to estimate SNP effects, and this can be done in either a univariate or multivariate manner. Univariate approaches attempt to estimate the marginal effect of a SNP on the phenotype and assess the statistical significance of SNP-phenotype associations while controlling for population structure using a genetic similarity matrix \cite{yu2006unified, kang2010variance, kang2008efficient}. However, as with all univariate testing approaches univariate LMM implementations may still produce spurious associations and biased effect estimates in the presence of LD.

Multivariate mixed model approaches reduce these problems by assessing the relationship between a phenotype of interest and all SNPs simultaneously via penalized regression, while controlling for environmental confounding  \cite{Rakitsch2012, bhatnagar2019simultaneous}. This approach assumes there are a relatively small number of SNPs associated with the phenotype of interest that have large effect sizes. These SNPs will be included in the model as both fixed and random effects, while SNPs with smaller effect sizes are modeled only as part of the random effect in order to control for environmental confounding.

The remainder of this review will focus on the performance of multivariate Lasso-penalized methods for the analysis of continuous, normally distributed outcomes. Specifically, PC-adjustment and LMM methods, which we will refer to as PC-Lasso and LMM-Lasso, respectively. We consider two similar implementations of the LMM-Lasso method, one developed by \cite{Rakitsch2012} and a more recent implementation from \cite{bhatnagar2019simultaneous}. Subsequent references to LMM-Lasso apply to both implementations, unless otherwise stated. Where necessary, we will differentiate between these implementations as LMM-Lasso-Rakitsch and LMM-Lasso-ggmix, respectively.

In section \ref{sec:results}, we compare the performance of PC-Lasso and LMM-Lasso in terms of their abilities to accurately estimate SNP effects and in terms of their true sign rates in the presence of varying levels of relatedness and environmental confounding. Although several studies have shown that LMMs outperform PC-adjustment methods in the univariate framework \cite{wang2013analytical, kang2010variance, zhao2007arabidopsis}, to the best of our knowledge this is the only comprehensive review and comparison of these methods in a penalized multivariate framework and using a simulation scheme motivated by human genetic relatedness and which formalizes the non-genetic confounding mechanism of population stratification.\\

@@

To describe and compare PC-Lasso and LMM-Lasso, we consider an $n \times p$ matrix of genotype data $\boldsymbol{G}$ for $n$ subjects and $p$ SNPs, where each $G_{i,j} \in \{ 0, 1, 2 \}$ enumerates the number of allele copies subject $i$ has for SNP $j$. We denote the normalized version of this genotype matrix as $\bX$, where each $X_{ij} = (G_{ij} - 2 p_j) / \sqrt{p_j (1 - p_j)}$ and $p_j$ is the minor allele frequency (MAF) of SNP $j$ \cite{zhang2015principal, price2006principal}. We assume that an $n \times 1$ vector of continuous phenotype values, $\by$, can be expressed as the sum of additive SNP effects, environmental confounding, and observational noise and consider the linear mixed model,
\begin{equation}
    \label{eqn:model}
    \by = \Int + \bX\bbeta + \bZ\bgamma + \beps
\end{equation}
where $\Int$ is the intercept, $\bbeta$ is a $p \times 1$ vector of SNP effects, $\bgamma$ is a $q \times 1$ vector of environmental effects, $\bZ$ is an $n \times q$ matrix that allocates environmental effects, and $\beps \sim N(\boldsymbol{0}, \sigmaee \bI_n)$. Often it is assumed that subjects belong to $q \le p$ subpopulations, and $\bZ$ is a corresponding incidence matrix of 0s and 1s. However, as we will discuss, this is not the only possibility.

LMM-Lasso relies on matrix decompositions of $\bX$ and the realized relationship matrix (RRM), $\bK$, which describes the pair-wise genetic similarity among subjects \cite{hayes2009increased}. Using the singular value decomposition (SVD) of the normalized genotype matrix, we can derive the spectral decomposition of the RRM:

@@

%------------------------------%
\subsection{PC-Lasso}
PC-Lasso uses the first $k$ principal components of $\bX$ to approximate $\bZ\bgamma$ and includes them in the model as unpenalized covariates. It has been shown \cite{hoffman2013correcting} that this corresponds to the regression model
\begin{equation}
    \label{eqn:pca_reg}
    \by = \Int + \bX\bbeta + \bU_{1:k}\boldsymbol{\alpha} + \beps 
\end{equation}
where $\bU_{1:k}$ is the $n \times k$ matrix of principal component vectors, $\boldsymbol{\alpha}$ their corresponding $k \times 1$ coefficient vector, and all other parameters are as defined in equation \eqref{eqn:model}. In a penalized regression framework, this model yields the objective function
\begin{equation}
    \label{eqn:pca_obj}
    Q(\boldsymbol{\beta, \alpha| X, y}) = \frac{1}{2n} ||\by - \Int - \bX \bbeta- \bU_{1:k}\boldsymbol{\alpha}||_2^2 + \lambda || \bbeta ||_1
\end{equation}
where $\lambda$ is a tuning parameter, and $|| \boldsymbol{\beta} ||_1 = \sum_{j=1}^p |\beta_j|$ denotes the $\ell_1$ norm of the regression coefficients, and $||.||_2$ the $\ell_2$ norm. Note that the principal component regression coefficients, $\boldsymbol{\alpha}$, are not included in the penalty term. Modeling unobserved confounding using fixed effects in this manner thus requires the estimation of $p + k + 1$ parameters.

%------------------------------%
\subsection{LMM-Lasso}
LMM-Lasso models $\bZ\bgamma$ as a random effect assuming 
$$\bgamma \sim N\left(\boldsymbol{0}, \frac{\sigmagg}{p} \bI_p \right).$$  Typically, $\bZ$ is not observed directly, and a common convention is to assume $\bZ = \bX$ \cite{wang2018multiplex, lippert2011fast, yang2014advantages}. Then, rather than assuming anything about $\bgamma$ directly, let $\bu=\bX\bgamma$ represent the random effect and assume $\bu \sim N(\boldsymbol{0}, \sigmagg \bK)$. This results in the LMM regression model
\begin{equation}
    \label{eqn:lmm_reg}
    \by = \Int + \bX\bbeta + \bX\bgamma+ \beps,
\end{equation}
where we assume $\beps \independent \bX\bgamma$ and thus $\by  \sim N(\Int + \bX \bbeta, \bV)$, where $\bV = \sigmagg \textbf{K} + \sigmaee \textbf{I}$. LMM-Lasso performs a weighted rotation of the original data, left-multiplying by $\bV^{-1/2}$ in a manner akin to generalized least squares to diagonalize the error term such that
\begin{equation}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}} \left(\Int +\bX \bbeta\right), \bI).
\end{equation}
\anna{...however, as Karl Rohe paper shows, GLS does not completely translate to penalized regression...comment on this?}
\begin{comment}
For ease of notation, we incorporate the intercept into the design matrix and let $\bX' = \begin{bmatrix} \textbf{1} & \bX \end{bmatrix}$ and $\bbeta' = \begin{pmatrix} \mu & \bbeta \end{pmatrix}^T$ so that we can express this more compactly as 
\begin{equation}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}} \bX' \bbeta', \bI).
\end{equation}
\end{comment}
Using the spectral decomposition of $\bK$ described in equation \eqref{eqn:k}, we can gain further insight into the mechanism of this transformation. Note that $\bV^{-1}$ can be expressed using the following matrix factorization 
\begin{align*}
    \bV^{-1} &= (\sigmagg \bK + \sigmaee \bI)^{-1}\\
    &=(\sigmagg \bU \bD^2 \bU^T + \sigmaee \bU \bU^T)^{-1}\\
    &= (\bU (\sigmagg \bD^2 + \sigmaee \bI) \bU^T)^{-1}\\
    &=\bU (\sigmagg \bD^2 + \sigmaee \bI)^{-1} \bU^T.
\end{align*}
By incorporating these decompositions, the transformed data can be expressed as $\bW \bXT$ and $\bW \byT$, respectively, where $\bXT = \bU^T \bX$ and $\byT = \bU^T \left( \by - \Int \right)$, where we center $\by$ to simplify notation. $\bW = \text{diag}\{w_1, ... w_n\}$ where $w_i$ corresponds to the square root of the $i$th element of the diagonal matrix $(\sigmagg \boldsymbol{D}^2 + \sigmaee \bI)^{-1}$. Together, this yields the objective function
\begin{equation}
\label{eqn:lmm_obj}
Q(\bbeta' | \bXT, \byT, \bW) = || \bW (\byT - \bXT \bbeta)||_2^2 + \lambda || \bbeta ||_1.
\end{equation}
An advantage to modeling unobserved confounding using a random effect, as opposed to fixed effects as in PC-Lasso, is that this requires the estimation of substantially fewer parameters: $p + 2$ for LMM-Lasso-Rakitsch, and $p + 3$ for LMM-Lasso-ggmix.

@@

We have so far used the term $\bW$ to describe the general LMM-Lasso method. However, the form of $\bW$ and method of variance component estimation differs between LMM-Lasso-Rakitsch and LMM-Lasso-ggmix. LMM-Lasso-Rakitsch \cite{Rakitsch2012} uses a two-step procedure for estimating the variance components, $\sigmaee$ and $\delta := \sigmaee / \sigmagg$. These variance components and the corresponding weights are estimated once under the assumption of a null model. A Lasso model can then be fit on the transformed data to estimate $\boldsymbol{\beta}$ using standard software for coordinate descent, such as \texttt{glmnet} \cite{glmnet} or \texttt{ncvreg} \cite{ncvreg}. LMM-Lasso-ggmix \cite{bhatnagar2019simultaneous} proposes iteratively updating $\widehat{\boldsymbol{\beta}}$ and the variance components via a block coordinate descent algorithm. This iterative update scheme necessitates the use of the corresponding \texttt{R} package \texttt{ggmix} \cite{ggmix, bhatnagar2019simultaneous}. We postpone further discussion of the differences between these methods to the results section.

>>>>

Genetic variants are often correlated with each other, meaning that univariate testing of genetic variants may result in false positives and biased effect estimates. This occurs when the variant being tested is correlated with other variants, so that the effects of these correlated variants are erroneously attributed to the variant in question. \anna{One method for analyzing data in the presence of correlated loci, or linkage disequilibrium (LD), is LD-pruning. In LD-pruning methods, SNPs are sequentially thinned, so that only SNPs correlated with each other below some desired threshold are used in modeling and testing procedures.} LD-pruning limits analysis to SNPs which are representative of a particular region. This may result in causal SNPs being removed from analysis, so that only SNPs in LD with the causal loci can be identified and their effects estimated. In addition, it has been shown that in the worst-case scenario, pruning algorithms may not select any representative SNPs in a particular region of the genome \citep{prive2018efficient}. However, LD-pruning does allow for the use of methods which assume uncorrelated loci such as \anna{BLAHHHHH: \url{https://www.nature.com/articles/srep27644/tables/1}}. In addition, LD-pruning may result in improved predictive accuracy \anna{because why}, which can be of particular benefit to polygenic risk score methods.

@@

The Lasso \citep{tibshirani1996regression} is a widely used penalized regression method. It relies on an assumption of sparsity, which means that coefficient estimation and variable selection can be accomplished simultaneously, and also provides considerable computational advantages, meaning that it scales up well to very large data sets. These features make the Lasso attractive for the analysis of high-dimensional genetic data where identifying a list of variants most highly associated with a particular trait is often of interest.

Lasso penalized regression models typically assume independent observations and uncorrelated errors. In the presence of population structure, some samples are more correlated than others, violating the independent errors assumption. If unaccounted for, this dependency can result in the identification of spurious associations. Though methods to correct for population stratification and cryptic relatedness have been extensively investigated \citep{Amin2007, hoffman2013correcting, price2006principal, Rakitsch2012, bhatnagar2019simultaneous, Sillanpaeae2011}, there remain open and relevant problems \pb{this is vague} in analyzing structured genetic data \citep{lawson2019population, barton2019population}.

One reason that population stratification warrants such concern in genetic studies is that due to ethnic and geographic segregation, population stratification tends to be associated with differing environmental exposures and cultural practices.  In other words, random genetic variation is likely to be associated with non-genetic factors that also influence the trait of interest, thereby confounding the genotype-phenotype relationship and leading to spurious associations and biased estimates of SNP effects. Biased estimates of this nature severely hinder prediction and understanding differences among populations \citep{barton2019population}. Although bias at individual loci may be small, this bias may become magnified when aggregated across thousands of SNPs, as is done when calculating polygenic risk scores \citep{barton2019population}.

In order to clarify and emphasize the importance of this subtle distinction, we present a detailed review of relevant concepts and methods. The remainder of this paper is organized as follows. In section \ref{sec:background} we provide a summary of the causes and consequences of structure in the genetic data of seemingly unrelated individuals and formally define terminology used throughout this paper. We emphasize that confounding due to population structure, frequently cited as a driver of spurious associations \citep{Sillanpaeae2011, sul2018population}, is a function of environmental heterogeneity, where genetic data serves as a proxy for differential environmental exposures. In so doing, we refine the definition of confounding due to population structure in terms of a non-genetic mechanism. In addition, we briefly review historical methods of correcting for population stratification and relatedness. In section \ref{sec:methods} we provide a detailed review of the two most prevalent multivariate methods of adjusting for population stratification and relatedness at present: PCA adjusted lasso penalized regression (PC-Lasso) and lasso penalized multivariate LMMs (LMM-Lasso). We review the statistical details of these methods as well as the consequences of the ways in which they attempt to correct for population structure and environmental heterogeneity. In section \ref{sec:results}, we illustrate the concepts described in section \ref{sec:methods} via simulation studies, where the data-generating mechanism formally differentiates between population structure and environmental heterogeneity. These simulation results are used to illustrate the importance of distinguishing between population structure and heterogeneous environmental exposures in penalized regression methods. In addition, these results identify scenarios where particular methods may be most and least effective in accurately estimating the effect sizes of observed SNPs in the presence of unobserved confounding.

@@

Population structure is a common phenomenon in genetic data. Varying levels of relatedness are almost always present among genetic samples, even in samples of unrelated individuals and seemingly homogeneous populations. For example, European American \citep{campbell2005demonstrating}, Han Chinese \citep{xu2009genomic, chen2009genetic}, and most recently, cohorts within the UK Biobank data \citep{haworth2019apparent}.

@@

Pedigree-based methods may be used to explicitly model recent relatedness if familial relationships are known. However, in this review we focus on methods to account for unobserved relatedness. Throughout this paper, the terms \textbf{population structure} and \textbf{structured population} will be used to describe a sample of individuals in which population stratification, cryptic relatedness, or both are present. Population stratification in particular has been of great concern in genetic studies due to its potential to lead to spurious associations. This phenomenon is commonly described as \textbf{confounding due to population stratification}. However, the mechanism of this phenomenon warrants some discussion. In particular, it is often overlooked that population structure itself does not confound the genotype-phenotype relationship in the classical sense, rather there must exist some non-genetic mechanism by which population stratification affects both the phenotype and genotype \citep{barton2019population, vilhjalmsson2012nature}. \\

@@

An additional consideration in genetic studies that we will briefly mention is that of genetic background. Genetic background describes when SNPs that have a causal relationship with the phenotype of interest are not directly observed, but are correlated with SNPs that are observed. As \citep{vilhjalmsson2012nature} note, any trait that has some genetic basis will be subject to genetic background effects, just as traits that are influenced by the environment may be subject to environmental confounding. Given the importance of both genetic and environmental contributions to many complex phenotypes, both genetic background and environmental confounding are important factors in our ability to identify true causal genetic relationships between SNPs and phenotype. Although there are a number of methods in which to account for genetic background \anna{what are they?}, the methods upon which we will focus address this by conducting a genomewide multivariate analysis.

@@

The genomic control (GC) method is a univariate testing procedure which modifies a test statistic by a deflation factor. This deflation factor attempts to quantify and remove the effect of population structure on the distribution of the appropriate test statistic, and is estimated using loci independent of the phenotype of interest and the loci being tested for association \citep{devlin1999genomic, bacanu2000power, wang2009testing}. However, this deflation factor has been reported to be sensitive to the nature and quantity of the loci used to estimate it \citep{hellwege2017population, marchini2004effects}. Additionally, GC does not correct effect size estimates, so resulting coefficient estimates will be unreliable after GC is applied, even if their corresponding test statistics and p-values have been appropriately adjusted \citep{hellwege2017population}.

Stratified analysis attempts to infer subject membership within a discrete number of nonoverlapping subpopulations. Subsequent analyses are conducted within each subpopulation, and the subpopulation-specific results may then be combined via meta-analysis \citep{pritchard1999use, pritchard2000association}. By partitioning the overall sample size in this manner, the structured association method is particularly vulnerable to a loss of power. Additionally, it relies on the assumption that these subpopulations are distinct and nonoveralapping. This assumption may not accurately reflect the characteristics of a particular sample, such as one consisting of admixed populations, or recent relatedness. Considered in the context of our air quality example, rather than the existence of two subpopulations with exposure to either good or bad air quality, there may also be individuals who fall between these two extremes and are characterized by continuously varying relatedness and a gradient of air quality. By clustering those individuals within in one subpopulation or another we lose information about their fine-grained, subtle relatedness.

Another approach is to create an approximation of population structure using surrogate variables, and to adjust for these as an additional model covariates. Principal component-adjustment (PC-adjustment) methods use the first several principal components of the design matrix as these surrogate variables \citep{price2006principal}. This approach may be used in univariate or multivariate frameworks, and allows for analysis on the entirety of a structured sample.

@@

Among LMM methods that aim to estimate heritability, the most well-known example is genome-wide complex trait analysis (GCTA) \citep{yang2011gcta}. In this framework, all observed SNPs are treated as random effects. These effects are integrated out in order to estimate a genetic variance component, which in turn, is used to quantify the total narrow-sense heritability of a trait \citep{yang2010common}.

LMMs can also be used to estimate SNP effects, and this can be done in either a univariate or multivariate manner. Univariate approaches attempt to estimate the marginal effect of a SNP on the phenotype and assess the statistical significance of SNP-phenotype associations while controlling for population structure using a genetic similarity matrix \citep{yu2006unified, kang2010variance, kang2008efficient}. However, as with all univariate testing approaches univariate LMM implementations may still produce spurious associations and biased effect estimates in the presence of LD.

Multivariate mixed model approaches reduce these problems by assessing the relationship between a phenotype of interest and all SNPs simultaneously via penalized regression, while controlling for environmental confounding  \citep{Rakitsch2012, bhatnagar2019simultaneous}. This approach assumes there are a relatively small number of SNPs associated with the phenotype of interest that have large effect sizes. These SNPs will be included in the model as both fixed and random effects, while SNPs with smaller effect sizes are modeled only as part of the random effect in order to control for environmental confounding.

The remainder of this review will focus on the performance of multivariate Lasso-penalized methods for the analysis of continuous, normally distributed outcomes. Specifically, PC-adjustment and LMM methods, which we will refer to as PC-Lasso and LMM-Lasso, respectively. We consider two similar implementations of the LMM-Lasso method, one developed by \citep{Rakitsch2012} and a more recent implementation from \citep{bhatnagar2019simultaneous}. Subsequent references to LMM-Lasso apply to both implementations, unless otherwise stated. Where necessary, we will differentiate between these implementations as LMM-Lasso-Rakitsch and LMM-Lasso-ggmix, respectively.

In section \ref{sec:results}, we compare the performance of PC-Lasso and LMM-Lasso in terms of their abilities to accurately estimate SNP effects and in terms of their true sign rates in the presence of varying levels of relatedness and environmental confounding. Although several studies have shown that LMMs outperform PC-adjustment methods in the univariate framework \citep{wang2013analytical, kang2010variance, zhao2007arabidopsis}, to the best of our knowledge this is the only comprehensive review and comparison of these methods in a penalized multivariate framework and using a simulation scheme motivated by human genetic relatedness and which formalizes the non-genetic confounding mechanism of population stratification.\\

@@

To describe and compare PC-Lasso and LMM-Lasso, we consider an $n \times p$ matrix of genotype data $\boldsymbol{G}$ for $n$ subjects and $p$ SNPs, where each $G_{i,j} \in \{ 0, 1, 2 \}$ enumerates the number of allele copies subject $i$ has for SNP $j$. We denote the normalized version of this genotype matrix as $\bX$, where each $X_{ij} = (G_{ij} - 2 p_j) / \sqrt{p_j (1 - p_j)}$ and $p_j$ is the minor allele frequency (MAF) of SNP $j$ \citep{zhang2015principal, price2006principal}. We assume that an $n \times 1$ vector of continuous phenotype values, $\by$, can be expressed as the sum of additive SNP effects, environmental confounding, and observational noise and consider the linear mixed model,
\begin{equation}
    \label{eqn:model}
    \by = \Int + \bX\bbeta + \bZ\bgamma + \beps
\end{equation}
where $\Int$ is the intercept, $\bbeta$ is a $p \times 1$ vector of SNP effects, $\bgamma$ is a $q \times 1$ vector of environmental effects, $\bZ$ is an $n \times q$ matrix that allocates environmental effects, and $\beps \sim N(\boldsymbol{0}, \sigmaee \bI_n)$. Often it is assumed that subjects belong to $q \le p$ subpopulations, and $\bZ$ is a corresponding incidence matrix of 0s and 1s. However, as we will discuss, this is not the only possibility.

LMM-Lasso relies on matrix decompositions of $\bX$ and the realized relationship matrix (RRM), $\bK$, which describes the pair-wise genetic similarity among subjects \citep{hayes2009increased}. Using the singular value decomposition (SVD) of the normalized genotype matrix, we can derive the spectral decomposition of the RRM:

@@

%------------------------------%
\subsection{PC-Lasso}
PC-Lasso uses the first $k$ principal components of $\bX$ to approximate $\bZ\bgamma$ and includes them in the model as unpenalized covariates. It has been shown \citep{hoffman2013correcting} that this corresponds to the regression model
\begin{equation}
    \label{eqn:pca_reg}
    \by = \Int + \bX\bbeta + \bU_{1:k}\boldsymbol{\alpha} + \beps 
\end{equation}
where $\bU_{1:k}$ is the $n \times k$ matrix of principal component vectors, $\boldsymbol{\alpha}$ their corresponding $k \times 1$ coefficient vector, and all other parameters are as defined in equation \eqref{eqn:model}. In a penalized regression framework, this model yields the objective function
\begin{equation}
    \label{eqn:pca_obj}
    Q(\boldsymbol{\beta, \alpha| X, y}) = \frac{1}{2n} ||\by - \Int - \bX \bbeta- \bU_{1:k}\boldsymbol{\alpha}||_2^2 + \lambda || \bbeta ||_1
\end{equation}
where $\lambda$ is a tuning parameter, and $|| \boldsymbol{\beta} ||_1 = \sum_{j=1}^p |\beta_j|$ denotes the $\ell_1$ norm of the regression coefficients, and $||.||_2$ the $\ell_2$ norm. Note that the principal component regression coefficients, $\boldsymbol{\alpha}$, are not included in the penalty term. Modeling unobserved confounding using fixed effects in this manner thus requires the estimation of $p + k + 1$ parameters.

%------------------------------%
\subsection{LMM-Lasso}
LMM-Lasso models $\bZ\bgamma$ as a random effect assuming 
$$\bgamma \sim N\left(\boldsymbol{0}, \frac{\sigmagg}{p} \bI_p \right).$$  Typically, $\bZ$ is not observed directly, and a common convention is to assume $\bZ = \bX$ \citep{wang2018multiplex, lippert2011fast, yang2014advantages}. Then, rather than assuming anything about $\bgamma$ directly, let $\bu=\bX\bgamma$ represent the random effect and assume $\bu \sim N(\boldsymbol{0}, \sigmagg \bK)$. This results in the LMM regression model
\begin{equation}
    \label{eqn:lmm_reg}
    \by = \Int + \bX\bbeta + \bX\bgamma+ \beps,
\end{equation}
where we assume $\beps \independent \bX\bgamma$ and thus $\by  \sim N(\Int + \bX \bbeta, \bV)$, where $\bV = \sigmagg \textbf{K} + \sigmaee \textbf{I}$. LMM-Lasso performs a weighted rotation of the original data, left-multiplying by $\bV^{-1/2}$ in a manner akin to generalized least squares to diagonalize the error term such that
\begin{equation}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}} \left(\Int +\bX \bbeta\right), \bI).
\end{equation}
\anna{...however, as Karl Rohe paper shows, GLS does not completely translate to penalized regression...comment on this?}
\begin{comment}
For ease of notation, we incorporate the intercept into the design matrix and let $\bX' = \begin{bmatrix} \textbf{1} & \bX \end{bmatrix}$ and $\bbeta' = \begin{pmatrix} \mu & \bbeta \end{pmatrix}^T$ so that we can express this more compactly as 
\begin{equation}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}} \bX' \bbeta', \bI).
\end{equation}
\end{comment}
Using the spectral decomposition of $\bK$ described in equation \eqref{eqn:k}, we can gain further insight into the mechanism of this transformation. Note that $\bV^{-1}$ can be expressed using the following matrix factorization 
\begin{align*}
    \bV^{-1} &= (\sigmagg \bK + \sigmaee \bI)^{-1}\\
    &=(\sigmagg \bU \bD^2 \bU^T + \sigmaee \bU \bU^T)^{-1}\\
    &= (\bU (\sigmagg \bD^2 + \sigmaee \bI) \bU^T)^{-1}\\
    &=\bU (\sigmagg \bD^2 + \sigmaee \bI)^{-1} \bU^T.
\end{align*}
By incorporating these decompositions, the transformed data can be expressed as $\bW \bXT$ and $\bW \byT$, respectively, where $\bXT = \bU^T \bX$ and $\byT = \bU^T \left( \by - \Int \right)$, where we center $\by$ to simplify notation. $\bW = \text{diag}\{w_1, ... w_n\}$ where $w_i$ corresponds to the square root of the $i$th element of the diagonal matrix $(\sigmagg \boldsymbol{D}^2 + \sigmaee \bI)^{-1}$. Together, this yields the objective function
\begin{equation}
\label{eqn:lmm_obj}
Q(\bbeta' | \bXT, \byT, \bW) = || \bW (\byT - \bXT \bbeta)||_2^2 + \lambda || \bbeta ||_1.
\end{equation}
An advantage to modeling unobserved confounding using a random effect, as opposed to fixed effects as in PC-Lasso, is that this requires the estimation of substantially fewer parameters: $p + 2$ for LMM-Lasso-Rakitsch, and $p + 3$ for LMM-Lasso-ggmix.

@@

We have so far used the term $\bW$ to describe the general LMM-Lasso method. However, the form of $\bW$ and method of variance component estimation differs between LMM-Lasso-Rakitsch and LMM-Lasso-ggmix. LMM-Lasso-Rakitsch \citep{Rakitsch2012} uses a two-step procedure for estimating the variance components, $\sigmaee$ and $\delta := \sigmaee / \sigmagg$. These variance components and the corresponding weights are estimated once under the assumption of a null model. A Lasso model can then be fit on the transformed data to estimate $\boldsymbol{\beta}$ using standard software for coordinate descent, such as \texttt{glmnet} \citep{glmnet} or \texttt{ncvreg} \citep{ncvreg}. LMM-Lasso-ggmix \citep{bhatnagar2019simultaneous} proposes iteratively updating $\widehat{\boldsymbol{\beta}}$ and the variance components via a block coordinate descent algorithm. This iterative update scheme necessitates the use of the corresponding \texttt{R} package \texttt{ggmix} \citep{ggmix, bhatnagar2019simultaneous}. We postpone further discussion of the differences between these methods to the results section.
